\documentclass[11pt]{article}

% ===> PACKAGES
\usepackage{amsmath,amsthm,amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{color}
\usepackage{comment}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage[margin=1in]{geometry} 
\usepackage{thmtools}

% ===> PARAMETERS
\pagestyle{fancy}

% ===> MACROS
\setlength{\parindent}{0em}
\setlength{\parskip}{0.5em}

%	\def\macrosName{Fill in the content of the macros and use \textit{\\macrosName} whenever necessary}
\newcommand\A{\mathcal{A}}
\newcommand\comp{\mathsf{comp}}
\newcommand\crash{\mathsf{crash}}
\newcommand\inbuf{\mathsf{inbuf}}
\newcommand\maj{\mathsf{maj}}

% Use these for theorems, lemmas, proofs, etc.
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{problem}{Problem}

\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
\DeclarePairedDelimiter\anglebrac{\langle}{\rangle}

\begin{document}

\lhead{CSC2221}
\chead{Xinyuan Li: 1000858244}
\rhead{\today}

\section*{Assignment 3}
% ===> START ASSIGNMENT
\begin{theorem}
\label{thm:byzantinenotpossible}
It is impossible to solve the Byzantine agreement deterministically in synchronous message passing network when up to $f$ processes can be Byzantine, if the network is not $(2f+1)$-connected.
\end{theorem}
\begin{proof}
Suppose for a contradiction that there exists an algorithm $\A$ which solves the Byzantine agreement problem deterministically in a synchronous message passing model when the network is not $(2f+1)$-connected. Consider the complete network on three processors where one processor is Byzantine. With $f = 1$ observe that this network is not $(2f+1)$-connected. By Theorem 5.7 of Attiya (stated as Theorem \ref{thm:Bthreeprocessors} here) no algorithm solves this consensus problem. This is a contradiction since $\A$ was suppose to be such an algorithm. \emph{If the question further requires that the that the number of processors $n > 3f$, then the complete graph on four nodes with one edge removed serves as a counter example. The proof is similar to the impossibility result on the triangle.}
\end{proof}

\begin{theorem}
\label{thm:Bthreeprocessors}
\textbf{(Theorem 5.7 in the Attiya Textbook.)} In a complete system with three processors and one Byzantine processor, there is no algorithm that solves the consensus problem.  
\end{theorem}

\begin{theorem}
\label{thm:byzantinepossible}
There exists a deterministic algorithm tolerating up to $f$ Byzantine processes that solves Byzantine agreement in a $(2f+1)$-connected network with $n > 3f$ processors.
\end{theorem}
\begin{proof}
We will describe such a deterministic algorithm and prove its correctness. This algorithm is inspired by the exponential bit complexity algorithm for Byzantine agreement in a complete network. Even though our network is no longer complete, we can simulate completeness by taking the majority of messages along node disjoint paths. \emph{We assume that global information about the graph is known to all processors; there is no indication in the question that this is not the case.}

Let $G = (V, E)$ be a $(2f+1)$-connected network where $|V| = n$, $|E| = m$, up to $f$ processors can be Byzantine, and $n > 3f$. Further, suppose each processor $p_i$ receives input $v_i$ and stores a private knowledge tree $K_i$. This tree will have the same structure as the tree for the exponential BA consensus algorithm in the complete network (i.e. on the first level of the tree there is a node associated with \emph{every} index $\anglebrac{i}$ where $1 \leq i\leq n$, on the second level there exists a node associated with every pair of indices $\anglebrac{i, j}$ for $1 \leq i, j\leq n$, etc.)

The algorithm is divided into two stages: first $p_i$ builds $K_i$ by sending requests along $(2f+1)$-node disjoint paths in $G$ and decorating $K_i$ with the appropriate information. Then $p_i$ analyzes the contents of $K_i$ by taking the recursive majority function starting at the root.   


(\textbf{Detailed description of the tree-building stage.}) $p_i$ receives a message from $p_j$ about level $l$ of $K_i$ as follows: $p_i$ decomposes $G$ into $(2f+1)$ internal-node-disjoint paths from $p_i$ to $p_j$. This is possible since $G$ is $(2f + 1)$-connected. For each path $t: p_i = p_{k_0} \leadsto p_{k_1} \leadsto \cdots \leadsto p_{k_c} = p_j$, $p_i$ sends message $\anglebrac{l, i=k_0, k_1, ..., k_c=j}$ to $p_{k_1}$. This message is passed along from $p_{k_l}$ to $p_{k_{l+1}}$ until it reaches $p_j$. If $p_j$ is a non-faulty processor it must return information about level $l$ of its knowledge tree back along $t$ to $p_i$. All processors will wait $2n$ rounds to deliver information about one level since this is the farthest distance between two processors. Once $p_i$ receives the $(2f+1)$ responses along all the node disjoint paths, it decorates the associated node in $K_i$ with the majority of the responses. This is the same procedure as for the complete network algorithm except for the $2n$ expansion in the number of rounds.

The analysis stage for each processor is identical to that of the BA consensus algorithm in a complete network. Each processor $p_i$ starts at the root and recursively takes the majority function. The majority of a leaf is simply the value of the leaf. The majority of any internal node is the majority of the values on its children. If a node does not have a majority or has an invalid value, its value is replaced with $\bot$ which is ignored by the majority function.

This algorithm terminates since the algorithm for BA consensus in a complete network terminates (this algorithm only takes $2n$ rounds longer). Validity and agreement both follow from validity and agreement of the algorithm for BA consensus in a complete network. These results transfer since the knowledge tree for each processor is decorated in the same way in both type of networks ($(2f+1)$-connected and complete). Since the paths chosen were node-disjoint and at most $f$ processors can be Byzantine, $f+1$ paths contain only non-faulty nodes. The results delivered on these paths form a majority and it is this majority value which decorates $K_i$ for processor $p_i$. 

\end{proof}
% ===> END ASSIGNMENT
\end{document}